WASP-4 is Accelerating Toward the Earth

### 2020-03-16 second_paper_review/response_20200315.txt

Dear Dr. Jackson,

We thank you for organizing the review of our paper, and thank the referee for
their helpful comments.  We have considered the referee's remarks carefully and
revised our manuscript accordingly.

Below, we have reproduced the referee's report, alongside with our
point-by-point responses. A list of changes is appended to the manuscript using
the trackchanges AASTeX macros.

Sincerely,

Luke Bouma

********************************************************************************
REFEREE COMMENT
----------
Section 2.1, Paragraph 2.

It would be helpful if the authors briefly describe the exact meaning of "high
quality" in the context of the transit times used in the analysis.

RESPONSE
----------
> Thank you for this prompt. We have removed the phrase "high quality", as its
> precise definition is not of particular relevance for this study. It is
> defined in Section 3.4 of Baluev+ 2019, through an empirical cutoff that
> Baluev+ made based on the sampling cadence and the light-curve residual RMS.

********************************************************************************
REFEREE COMMENT
----------
Section 3.1

In this section it is stated that the assumption of "a Gaussian likelihood and
sampling over the posterior probability distributions" is made. This assumption
needs further justification given the variety of data sources included. In
particular, it is later stated in Section 3.1 that the cause of the high
reduced chi2 values "must be some additional signal or noise". Alternatively,
it could indeed mean that the Gaussian likelihood assumption is not completely
valid.

RESPONSE
----------
> TODO

----------
********************************************************************************
REFEREE COMMENT
----------
Section 3.3

The discussion and figure in this section are very good. However, it is not
sufficiently demonstrated that a stellar companion is ruled out when
considering the possible angular separations of the companion resulting from
being near inferior or superior conjunction. This is particularly important for
this system since even a small degree of coplanarity of the companion with
WASP-4b would result in an edge on orbit that makes the companion difficult to
detect when near conjunction locations. Also, the very last sentence of this
section "Figure 4 shows the result." should describe the results of the figure
in more detail.

RESPONSE
----------
> TODO
>
> The comment regarding the last sentence has been addressed.

********************************************************************************
REFEREE COMMENT
----------
Section 4.1

Paragraph 1. The paragraph implies that the Doppler effect is the sole cause of
the timing derivative. However, and given the 2 sigma discrepancy, the wording
should allow for the possibility that the timing derivative is actually a
result of a combination of the described effects.

RESPONSE
----------
> We have modified the wording in this paragraph to clarify that this is a
> possibility, but that at present the data are best explained

********************************************************************************
REFEREE COMMENT
----------
Section 4.2

Paragraph 3. "hot Jupiters is expected" should be "hot Jupiters are expected".

RESPONSE
----------
> Corrected, thank you.

### 2020-03-16 paper/ms.tex
[WIP] second referee report

\documentclass[12pt,twocolumn,tighten]{aastex62}
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}
\usepackage{natbib}

@@

Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.  Additional timing measurements were also
reported recently by \citet{baluev_2019}, based on a homogeneous
analysis of archival ground-based observations.  We included twelve of
their ``high quality'' transit times from the TRAPPIST telescope (six
transits), the El~Sauce 36$\,$cm (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that the timestamps were amenable to the
appropriate barycentric and leap second corrections (M.~Gillon,
P.~Evans, priv{.} comm{.}).  We omitted the fourteen remaining
ETD\footnote{\url{http://var2.astro.cz/ETD}} times from
\citeauthor{baluev_2019} due to ambiguity in whether leap-second
corrections had been performed.  We did not include in our analysis
the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

The resulting period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within 1-$\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is 2.3-$\sigma$ larger than the rate of
period decrease reported by \citetalias{bouma_wasp4b_2019}
($- 12.6 \pm 1.2~{\rm ms}\,{\rm yr}^{-1}$), presumably because of the new
data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}. The other best-fitting model parameters are
reported in Table~3.

@@

For each simulated companion, we drew a set of model parameters from
the converged posterior probability samples of our model for the
WASP-4 radial velocities.  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis.
Figure~\ref{fig:mass_sma} shows the result.

@@

The previously offered explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, apsidal precession, and the
Doppler effect \citep{bouma_wasp4b_2019}.  Our new radial velocity
measurements strongly indicate that the least exotic option---the
Doppler effect---is the most likely explanation.  Long-term transit
timing data show the orbital period to be decreasing by $- 8.64 \pm
1.26~{\rm ms}\,{\rm yr}^{-1}$.  The long-term radial velocity data
show a trend that should lead to an apparent period derivative of
$-5.94 \pm 0.39~{\rm ms}\,{\rm yr}^{-1}$ through the Doppler effect.
Although these two measurements of the period derivative are
discrepant by about 2-$\sigma$, Occam's razor suggests that most or
all of the apparent decrease of WASP-4b's orbital period is caused by
the line-of-sight acceleration.

@@

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
Therefore, around 1 in 3 hot Jupiters is expected to show period
changes comparable to that of WASP-4 due to acceleration by outer
companions.  The sign of the accelerations should be random, with
about half of them approaching and half of them receding.  With a
large enough sample of short-period hot Jupiters, one might be able to
distinguish line-of-sight accelerations from tidal orbital decay by
seeking evidence for a systematic tendency for the observed period to
decrease, rather than increase.

@@

>>>>

%\documentclass[12pt,twocolumn,tighten]{aastex62}
\documentclass[12pt,twocolumn,tighten,trackchanges]{aastex62}
\usepackage{amsmath,amstext,amssymb}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage[figure,figure*]{hypcap}
\usepackage{graphics,graphicx}
\usepackage{hyperref}
\usepackage{natbib}

@@

Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.  Additional timing measurements were also
reported recently by \citet{baluev_2019}, based on a homogeneous
analysis of archival ground-based observations.  We included
\deleted{twelve of} their \deleted{``high quality''} transit times
from the TRAPPIST telescope (six transits), the El~Sauce 36$\,$cm
(four transits), and \citet{petrucci_no_2013} \added{(two transits)}.
For TRAPPIST and El~Sauce, we verified with the original observers
that the timestamps were amenable to the appropriate barycentric and
leap second corrections (M.~Gillon, P.~Evans, priv{.} comm{.}).  We
omitted the fourteen remaining
ETD\footnote{\url{http://var2.astro.cz/ETD}} times from
\citeauthor{baluev_2019} due to ambiguity in whether leap-second
corrections had been performed.  We did not include in our analysis
the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

The resulting period derivative for the quadratic model is 
\begin{equation}
\dot{P}
  = - (2.74 \pm 0.28)\times 10^{-10}
  = - 8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}.
  \label{eq:dP_dt_obs}
\end{equation}
This agrees to within 1-$\sigma$ of the value reported by
\citet{southworth_transit_2019} ($\dot{P} = - 9.2 \pm 1.1~{\rm
ms}\,{\rm yr}^{-1}$).  It is 2.3-$\sigma$ larger than the rate of
period decrease reported by \citetalias{bouma_wasp4b_2019}
($- 12.6 \pm 1.2~{\rm msec}\,{\rm yr}^{-1}$), presumably because of the new
data from \citeauthor{southworth_transit_2019} and
\citeauthor{baluev_2019}. The other best-fitting model parameters are
reported in Table~3.

@@

For each simulated companion, we drew a set of model parameters from
the converged posterior probability samples of our model for the
WASP-4 radial velocities.  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis.
Figure~\ref{fig:mass_sma} shows the result\replaced{.}{: the companion
responsible for the acceleration has a true mass of 10-200$\,M_{\rm
Jup}$ and an orbital distance of $10$-$100\,{\rm AU}$.}

@@

The previously offered explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, apsidal precession, and the
Doppler effect \citep{bouma_wasp4b_2019}.  Our new radial velocity
measurements strongly indicate that the least exotic option---the
Doppler effect---is the \replaced{most likely explanation}{dominant
physical process}.  Long-term transit timing data show the orbital
period to be decreasing by $- 8.64 \pm 1.26~{\rm msec}\,{\rm
yr}^{-1}$.  The long-term radial velocity data show a trend that
should lead to an apparent period derivative of $-5.94 \pm 0.39~{\rm
msec}\,{\rm yr}^{-1}$ through the Doppler effect.  Although these two
measurements of the period derivative are discrepant by about
2-$\sigma$, Occam's razor \replaced{suggests that most or all of}{would suggest that} the
apparent decrease of WASP-4b's orbital period is caused
\added{solely} by the line-of-sight acceleration.\added{ Detection
of any additional second-order effects will require more precise data,
and a more significant discrepancy.}

@@

Including WASP-4b, 16 of 51 hot Jupiters surveyed by
\citet{knutson_friends_2014} show a non-zero radial velocity trend.
Therefore, around 1 in 3 hot Jupiters \replaced{is}{are} expected to
show period changes comparable to that of WASP-4 due to acceleration
by outer companions.  The sign of the accelerations should be random,
with about half of them approaching and half of them receding.  With a
large enough sample of short-period hot Jupiters, one might be able to
distinguish line-of-sight accelerations from tidal orbital decay by
seeking evidence for a systematic tendency for the observed period to
decrease, rather than increase.

@@

\listofchanges
